---
title: "Lab 2b: Passive Imputation"
subtitle: "Missing Data in R"
author: "Kyle M. Lang"
date: "Updated: `r format(Sys.time(), format = '%Y-%m-%d')`"
params:
  answers: false
output: 
   bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: true
    css: "../../resources/style.css"
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)

set.seed(235711)

## Define an asis engine that will evaluate inline code within an asis block:
knit_engines$set(asis = function(options) {
  if(options$echo && options$eval) knit_child(text = options$code)
}
)

opts_chunk$set(include = params$answers, 
               echo = params$answer, 
               message = FALSE,
               warning = FALSE,
               fig.align = "center")
```

---

In these lab exercises, you will explore how certain pathological data 
characteristics can affect imputation results. You will also learn a very useful 
technique for imputing variables with known relations: *Passive Imputation*.

---

# Setup

---

##

**Load the `mice` package.**

```{r}
library(mice)
```

---

The *mammalsleep* dataset is part of **mice**. This dataset contains the 
[Allison and Cicchetti (1976)][data_ref] data for mammalian species.

- Unless otherwise specified, all questions in this section refer to the 
*mammalsleep* data.

---

##

**Check the documentation for the *mammalsleep* dataset.**

```{r, eval = FALSE}
?mammalsleep
```

---

##

**Summarize the data to get an overview of their structure.**

```{r}
head(mammalsleep)
summary(mammalsleep)
str(mammalsleep)
```

---

##

**Compute the missing data patterns in the data.**

- How many different missing data patterns are present in the data?
- Which pattern occurs most frequently in the data?

```{r}
(pats <- md.pattern(mammalsleep))
```

```{asis}
The *mammalsleep* data contain `r nrow(pats) - 1 ` total response patterns. The 
complete case pattern (i.e., the pattern wherein all variables were observed) 
occurred the most frequently (`r rownames(pats) %>% as.numeric() %>% max(na.rm = TRUE)` 
times).
```

---

# Naive Imputation

---

## {#imp1}

**Use `mice()` to multiply impute the *mammalsleep* data. **

- Create 5 imputations.
- Use 10 iterations.
- Use predictive mean matching as the imputation method.
- Set a random number seed.
- Keep all other options at their default values.

```{r, cache = TRUE}
imp <- mice(mammalsleep, maxit = 10, seed = 235711, print = FALSE)
```

When you run this imputation, you will probably see a warning about "logged 
events". When `mice()` encounters certain computational difficulties (e.g., 
extreme collinearity), it will take automatic remedial action and log the action 
in the `loggedEvents` slot of the *mids* object.

If you ever see a warning about logged events, you should check the `loggedEvents` 
slot to see what actions were taken. You want to assess if the actions were 
appropriate and judge the likely impact that the actions will have on your 
results. If the actions `mice()` has taken seem too extreme, you need to address 
the underlying data issues and rerun the imputation with the cleaned data.

---

##

**Check the contents of the `loggedEvents` slot in the *mids* object you created 
in Question \@ref(imp1).** 

- What problem is `mice()` logging here?

```{r}
imp$loggedEvents %>% head()
```

```{asis}
The imputation models seem to have more predictors ($P = 71$) than observations 
($N \approx 50$). It looks like something may have gone "wrong" with the `species` 
variable. To get around the issue, `mice()` has applied a ridge penalty.
```

---

We can get a sense of how impactful the issues noted in the `loggedEvents` log 
have been through our usual diagnostic plots.

---

##

**Create trace plots, density plots, and strip plots from the *mids* object you 
created in Question \@ref(imp1).**

What do you conclude vis-a-vis convergence and the validity of these imputations?

```{r}
plot(imp)
densityplot(imp)
stripplot(imp)
```

```{asis}
The trace plots look OK. So, it appears that the imputation model have converged 
onto some equilibrium. The density plots and the strip plots, however, suggest 
very poor imputations. The density plots, in particular, clearly show far too 
little variability in the imputed values. Nearly all of the imputations collapse 
onto a small range of values (hence the "sharp" spikes in the density plots). 
Although the imputation models may have converged, it appears they have converged 
onto the wrong solution. These imputations are certainly not reasonable.
```

---

For didactic purposes, let's "play the fool", ignore any information we may have 
gleaned from the diagnostic plots, and analyze these imputed data via the usual 
process. 

---

##

**Use the imputed data you created in Question \@ref(imp1) to fit the following 
regression model.**

$Y_{sws} = \beta_0 + \beta_1 X_{ln(bw)} + \beta_2 X_{odi} + \varepsilon$

Pool the MI estimates, and check the RIV, $\lambda$, and FMI.

- What do you notice?

```{r}
est <- with(imp, lm(sws ~ log(bw) + odi)) %>% pool()
summary(est)
est
```

```{asis}
Although the estimates seem sensible, the RIV, $\lambda$, and FMI values all 
suggest that the missing data have had a very large influence on the results. 
For example, the $\lambda$ for $\hat{\beta}_1$ tells us that 
`r 100 * est$pooled$lambda[2] %>% round(1)`% of the sampling variance in 
$\hat{\beta}_1$ is attributable to the missing data and our treatment thereof. 
Although we already knew these imputations were suspect, these regression 
results further confirm the poor quality of the imputations.
```

---

The poor performance you should have noted above is largely driven by the 
`species` variable. This variable is a factor with `r nlevels(mammalsleep$species)` 
levels. So, when we include this variable as a predictor in the imputation models, 
it enters the model as a set of `r nlevels(mammalsleep$species) - 1` dummy codes. 
These dummy codes produce the $P > N$ problem noted in the `loggedEvents` log 
which leads to poor imputations.

---

## {#imp2}

**Use `mice()` to re-impute the *mammalsleep* data.**

Use the same settings from Question \@ref(imp1), but do not use `species` as a 
predictor in any of the imputation models.

- Name the resulting *mids* object `imp1`.

```{r, cache = TRUE}
pred <- imp$predictorMatrix
pred[ , "species"] <- 0

imp1 <- mice(mammalsleep, 
             predictorMatrix = pred, 
             maxit = 10, 
             seed = 235711, 
             print = FALSE)
```

##

**If you get a warning about logged events. Check the `loggedEvents` slot of the 
*mids* object you created in Question \@ref(imp2).**

```{r}
imp1$loggedEvents
```

```{asis}
This time, the logged events are telling us about collinearity problems and the 
actions taken to remedy the collinearity. Specifically, when imputing `mls` and 
`gt`, `ts` was collinear with other predictors, so it was removed from the model.
```

---

##

**Create trace plots, density plots, and strip plots from the *mids* object you 
created in Question \@ref(imp2).**

What do you conclude vis-a-vis convergence and the validity of these imputations?

```{r}
plot(imp1)
densityplot(imp1)
stripplot(imp1)
```

```{asis}
This time, the trace plots suggest some serious identification issues. Notice 
how the individual lines in the plots of the means for `ps` and `ts` are stable 
but do not mix. This pattern is indicative of an under-identified model. 
Basically, the data do not contain enough information to define a unique
solution for some parameters.

The imputations look pretty much fine, but we don't care. The imputation model 
must converge before we can move on to considering the plausibility of the 
imputed values.
```

---

The convergence issues you should have noticed above are caused by structural 
features of the data. Total sleep (`ts`) is the sum of paradoxical sleep (`ps`) 
and short wave sleep (`sws`). The imputation model treats `ts` as distinct and 
stochastically related to `ps` and `sws`, but `ts` is actually a deterministic 
function of `ps` and `sws`. This deterministic relation is ignored in the 
imputations, and the resulting circularity in the imputations keeps the model 
from finding a unique solution. 

Thankfully, `mice()` offers a convenient routine for addressing exactly these 
types of known relations among variables in the imputation model: *passive 
imputation*. With passive imputation, we can account for transformations, 
combinations, and recoded variables when imputing their missing data. 

---

# Passive Imputation

---

Frequently, we need to transform, combine, or recode variables. When such a need 
arises with incomplete variables that we'd like to impute, we have a few options.

1. We could impute the original, un-transformed, variable and transform the 
completed version afterwards (the so-called *impute-then-transform* approach).
1. We could transform the incomplete variable and impute the transformed version 
as if it were any other variable (the so-called *just-another-variable* approach). 

Both of these approaches have an important limitation, though. In neither case 
does the imputation model have access to both the original and the transformed 
versions of the variable in question. The imputations are either generated using 
the information in the raw version of the variable (in the impute-then-transform 
approach) or using the information in the transformation (in the 
just-another-variable approach), but not both. Note that keeping both the raw 
and transformed versions of a variable in the model is not an option since doing 
so induces perfectly collinear variables. 

To solve this problem, `mice()` implements a third approach called *passive 
imputation*. The goal of passive imputation is to maintain known, deterministic 
relations among incomplete variables throughout the imputation process and to 
allow the imputation model to use the transformed variables as predictors when 
imputing other variables (other than the raw version of the transformed variables, 
themselves).

For example, we can use passive imputation to maintain the following 
deterministic function in the `boys` data
$$\text{BMI} = \frac{\text{Weight}}{\text{Height}^2}$$
or this compositional relation in the `mammalsleep` data
$$\text{ts} = \text{ps}+\text{sws}.$$

---

To implement passive imputation, we need to adjust two features of the `mice()` setup:

1. The method vector
   - We use the method vector to define the deterministic relations.
1. The predictor matrix
   - We adjust the predictor matrix to keep a transformed variable from being 
   used as a predictor of its raw version.

---

The following code will adjust the method vector from Question \@ref(imp2) to 
implement passive imputation for `ts`.

```{r, include = TRUE, echo = TRUE}
(meth <- imp1$method)
meth["ts"]<- "~ I(sws + ps)"
meth
```

Now, `ts` will not be independently imputed along with the other variables. 
Rather, in each iteration, the most recently completed version of `sws` and `ps` 
will be added together to define the updated version of `ts`. 

---

The updated version of `ts` defined according to the deterministic relation 
described above can then be used as a predictor when imputing other variables, 
but we do not want to use `ts` to impute either `sws` or `ps` (to avoid 
circularity). So, we need to adjust the predictor matrix to satisfy this 
restriction. 

```{r, include = TRUE, echo = TRUE}
(pred <- imp1$predictorMatrix)
pred[c("sws", "ps"), "ts"] <- 0
pred
```

Now, we can re-impute the *mammalsleep* data using passive imputation to account 
for the deterministic relation between `ts`, `sws`, and `ps`. We do so simply by 
using the updated method vector and predictor matrix in a regular run of `mice()`. 

```{r, include = TRUE, echo = TRUE, cache = TRUE}
imp <- mice(mammalsleep, 
            method = meth, 
            predictorMatrix = pred, 
            maxit = 10, 
            seed = 235711, 
            print = FALSE)
```

---

If we inspect the diagnostic plots for these imputations, we see much better 
performance than we achieved in Questions \@ref(imp1) or \@ref(imp2).

```{r, include = TRUE, echo = TRUE}
plot(imp)
densityplot(imp)
stripplot(imp)
```

We can see that the pathological non-convergence of Question \@ref(imp2) has 
been corrected by the passive imputation. 

---

You will now implement passive imputation yourself using the *boys* dataset from 
the **mice** package. You already explored the missing data in the *boys* data 
in the last practical, so we will not repeat those steps. 

Unless otherwise specified, all remaining questions refer to the *boys* dataset.

---

## {#passImp1}

**Multiply impute the *boys* data using passive imputation for `bmi`.**

Use passive imputation to maintain the known relation between `bmi`, `wgt`, and 
`hgt`.

- Specify the method vector entry for `bmi` as `"~ I(wgt / (hgt / 100)^2)"`.
- Keep the default predictor matrix, for now.
- Use 20 iterations.
- Set a random number seed.
- Leave all other settings at there default values.
- Name the resulting *mids* object `imp1`.

```{r, cache = TRUE}
## Use the mice::make.method() function to generate a default method vector:
(meth <- make.method(boys))

meth["bmi"] <- "~ I(wgt / (hgt / 100)^2)"
meth

imp1 <- mice(boys, method = meth, maxit = 20, seed = 235711, print = FALSE)
```

---

Run the following code to inspect the relation between the imputed BMI and the 
BMI calculated from the imputed height and weight. If the passive imputation was 
successful, these points should fall along a perfect line.

```{r, include = TRUE, echo = TRUE}
xyplot(imp1, 
       bmi ~ I(wgt / (hgt / 100)^2), 
       ylab = "Imputed BMI", 
       xlab = "Calculated BMI")
```

---

##

**Create trace plots, density plots, and strip plots from the *mids* object you 
created in Question \@ref(passImp1).**

What do you conclude vis-a-vis convergence and the validity of these imputations?

```{r}
plot(imp1)
densityplot(imp1)
stripplot(imp1)
```

```{asis}
Although the deterministic definition of `bmi` is now preserved in the completed
data, the trace plots indicate some pathological behavior for `bmi`, `hgt`, and 
`wgt`. We also get some absurd imputations for `bmi`. 
```

---

Of course, the issues you should have spotted in the above imputations are to 
be expected since we have purposefully omitted the second part of passive 
imputation. We have not adjusted the predictor matrix, so we have circularity in 
the imputations. We used passive imputation to create the imputations for `bmi`, 
but `bmi` is still used as predictor for `wgt` and `hgt`. 

---

## {#passImp2}

**Adjust the predictor matrix to remove the circularity described above.**

Re-impute the *boys* data using the updated predictor matrix.

- Keep all other settings the same as in Question \@ref(passImp1).
- Name the resulting *mids* object `imp1`.

```{r, cache = TRUE}
(pred <- imp1$predictorMatrix)
pred[c("hgt", "wgt"), "bmi"] <- 0
pred

imp1 <- mice(boys, 
             method = meth, 
             predictorMatrix = pred, 
             maxit = 20,
             seed = 235711, 
             print = FALSE)
```

---

##

**Recreate the `xyplot()` from above using the imputations from Question 
\@ref(passImp2).**

Is the deterministic definition of `bmi` maintained in the imputed data?

```{r}
xyplot(imp1, 
       bmi ~ I(wgt / (hgt / 100)^2), 
       ylab="Imputed BMI", 
       xlab="Calculated BMI")
```

```{asis}
Yes, the relation is maintained. All points fall along the $Y = X$ line.
```

---

##

**Create trace plots, density plots, and strip plots from the *mids* object you 
created in Question \@ref(passImp2).**

What do you conclude vis-a-vis convergence and the validity of these imputations?

```{r}
plot(imp1)
densityplot(imp1)
stripplot(imp1)
```

```{asis}
Everything looks good now. The trace plots indicate good convergence (though we 
clearly need more than the default 5 iterations for the model to stabilize). 
Judging from the density plots and strip plots, the imputations also seem sensible.
```

---

**Just for fun: What you shouldn’t do with passive imputation**

Never fix all relations. The algorithm will never escape the starting values.

```{r, include = TRUE, echo = TRUE, cache = TRUE}
meth <- make.method(boys)

meth["bmi"] <- "~ I(wgt / (hgt / 100)^2)"
meth["wgt"] <- "~ I(bmi * (hgt / 100)^2)"
meth["hgt"] <- "~ I(sqrt(wgt / bmi) * 100)"

imp <- mice(boys, method = meth, seed = 235711, print = FALSE)

plot(imp, c("hgt", "wgt", "bmi"))
```

---

End of Lab 2b

---
[data_ref]: http://doi.org/10.1126/science.982039
[rubin_1996]: http://doi.org/10.1080/01621459.1996.10476908
[wu_et_al_2015]: https://doi.org/10.1080/00273171.2015.1022644
